---
title: "Analysis of BLAST Results"
author: "Don Francisco"
date: "October 6, 2017"
output: github_document
---

# Introduction

Add 1-2 paragraphs here.

# Methods

## Sample origin and sequencing

Add about a paragraph here.

## Computational

And another paragraph or two here.

# Results

```{r load-libraries, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
```

```{r make-read-in-data-function}
# Output format from BLAST is as detailed on:
# https://www.ncbi.nlm.nih.gov/books/NBK279675/
# In this case, we used: '10 sscinames std'
# 10 means csv format
# sscinames means unique Subject Scientific Name(s), separated by a ';'
# std means the standard set of result columns, which are:
# 'qseqid sseqid pident length mismatch
# gapopen qstart qend sstart send evalue bitscore',


# this function takes as input a quoted path to a BLAST result file
# and produces as output a dataframe with proper column headers
# and the 'qseqid' column split into sample and seq number
read_blast_output <- function(filename) {
  data_in <- read.csv(filename,
                      header = FALSE, # files don't have column names in them
                      col.names = c("sscinames", # unique Subject Sci Name(s)
                                    "qseqid",    # Query Seq-id
                                    "sseqid",    # Subject Seq-id
                                    "pident",    # Percntge of identical matches
                                    "length",    # Alignment length
                                    "mismatch",  # Number of mismatches
                                    "gapopen",   # Number of gap openings
                                    "qstart",    # Start of alignment in query
                                    "qend",      # End of alignment in query
                                    "sstart",    # Start of alignment in subj
                                    "send",      # End of alignment in subject
                                    "evalue",    # Expect value
                                    "bitscore"))  # Bit score

  # Next we want to split the query sequence ID into
  # Sample and Number components so we can group by sample
  # They originally look like "ERR1942280.1"
  # and we want to split that into two columns: "ERR1942280" and "1"
  # we can use the separate() function from the tidyr library to do this
  # Note that we have to double escape the period for this to work
  # the syntax is
  # separate(column_to_separate,
  # c("New_column_name_1", "New_column_name_2"),
  # "seperator")
  data_in <- data_in %>%
    separate(qseqid, c("sample_name", "sample_number"), "\\.")
}
```

```{r read-in-BLAST-data}
# this makes a vector of all the BLAST output file names, including
# the name(s) of the directories they are in
files_to_read_in <- list.files(path = "output/blast",
                               full.names = TRUE)

# We need to create an empty matrix with the right number of columns
# so that we can rbind() each dataset on to it
joined_blast_data <- matrix(nrow = 0,
                            ncol = 14)

# now we loop over each of the files in the list and append them
# to the bottom of the 'joined_blast_data' object
# we do this with the rbind() function and the function we
# made earlier to read in the files, read_blast_output()
for (filename in files_to_read_in) {
  joined_blast_data <- rbind(joined_blast_data,
                             read_blast_output(filename))
}
```

```{r read-in-metadata-and-join}
# Next we want to read in the metadata file so we can add that in too
# This is not a csv file, so we have to use a slightly different syntax
# here the `sep = "\t"` tells the function that the data are tab-delimited
# and the `stringsAsFactors = FALSE` tells it not to assume that things are
# categorical variables
metadata_in <- read.table(paste0("data/metadata/",
                                 "fierer_forensic_hand_mouse_SraRunTable.txt"),
                          sep = "\t",
                          header = TRUE,
                          stringsAsFactors = FALSE)

# Finally we use the left_join() function from dplyr to merge or 'join' the
# combined data and metadata into one big table, so it's easier to work with
# in R the `by = c("Run_s" = "sample_name")` syntax tells R which columns
# to match up when joining the datasets together
joined_blast_data_metadata <- metadata_in %>%
  left_join(joined_blast_data,
            by = c("Run_s" = "sample_name"))
```


```{r histograms}
# Here we're using the dply piping syntax to select a subset of rows matching a
# criteria we specify (using the filter) function, and then pull out a column
# from the data to make a histogram. We don't need to tell the hist() function
# which data to use, because that's piped in, but we do have to give the
# hist() function the title and axis label we'd like to use for the figure
joined_blast_data_metadata %>%
  filter(env_material_s == "sebum") %>%
  pull(pident) %>%
  hist(main = "Percent Identity",
       xlab = "Percent",
       col = "lightcoral")
```

Figure 1. Histogram showing the percent identity matches of sebum in male and female subjects.

```{r Dust-on-Computer Mouse}
joined_blast_data_metadata %>%
filter(env_material_s == "dust") %>%
filter(sample_type_s == "computer mouse") %>%
  pull(pident) %>%
  hist(main = "Percent Dust Found on Computer Mice",
       xlab = "Percent",
       col = "forest green") 
```

Figure 2. This histogram shows the bacterial percent identity in dust found on computer mice. 

```{r histogram-female-sebum}
joined_blast_data_metadata %>%
filter(env_material_s == "sebum") %>%
filter(sex_s == "female") %>%
  pull(pident) %>%
  hist(main = "Female Percent Identity of Sebum",
       xlab = "Percent",
       col = "plum")
```

Figure 3. This graph shows that between the percent bacterial identity of sebum found on both genders, females palms exhibited a very high percentage.

```{r histogram-male-sebum}
joined_blast_data_metadata %>%
filter(env_material_s == "sebum") %>%  
filter(sex_s == "male") %>%
  pull(pident) %>%
  hist(main = "Male Percent Identity of Sebum",
       xlab = "Percent",
       col = "skyblue")
```

Figure 4. This graph shows that among the bacterial identities in the sebum samples taken, male hands had a significantly lower match than that of females. 

```{r sebum-sequence-length}
joined_blast_data_metadata %>%
filter(env_material_s == "sebum") %>%
filter(length > 150) %>%
  pull(length) %>%
  hist(main = "Sebum Sequence Length",
       xlab = "sequence length",
       col = "orange")
```

Figure 5. This graph shows the sequence length matches (in bp) on both males and female for sequence lengths over 150.

```{r table-sequence-length}


```{r intro-to-ggplot}
ggplot(joined_blast_data_metadata,
       aes(x = sex_s,
           y = length)) +
  geom_boxplot()
```

```{r summary-table}
joined_blast_data_metadata %>%
  group_by(sscinames) %>%
  count() %>%
  arrange(desc(n)) %>%
  head(10) %>% 
  kable(digits = 5)
```

Table 1: Table 1 showing the top 10 bacterial identity matches and their corresponding sequence lengths

# Finally, we'd like to be able to make a summary table of the counts of
# sequences for each taxa for each sample. To do that we can use the table()
# function. We add the kable() function as well (from the tidyr package)
# in order to format the table nicely when the document is knitted
kable(table(joined_blast_data_metadata$sscinames,
            joined_blast_data_metadata$Run_s))
```

# Discussion

Add 2-3 paragraphs here interpreting your results and considering future directions one might take in analyzing these data.


